---
title: "Maps and Data Viz"
author: "Seamus Stein"
date: "11/4/2023"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Inputs: 
final_data.csv: The csv file containing all of the joined data for the regression analysis 
Africa_Boundries.csv: a csv file containing ISO codes for each country in Africa. 

ISO codes are inner joined to the final data csv file (read in as malaria_data) for mapping in plotly.

Required Packages:
tidyverse
magrittr
plotly
shiny
ggcorrplot

Output: 
A shiny dashboard, with 8 maps of Africa. For 2019 and 2020 respectively there is a map of log transformed cases, log transformed deaths, malaria prevalence and malaria mortality rate. 

A correlation matrix using Spearman's correlation is also created and displayed at the end. 

```{r}
library(tidyverse)
library(magrittr)
library(plotly)
library(shiny)
library(ggcorrplot)
```


```{r}
# read in the data
malaria_data <- malaria <-read_csv("~/Desktop/project_data/final_data.csv")
iso_codes <- read_csv("~/Desktop/project_data/Africa_Boundaries.csv")
```

```{r}
# rename name of country to Location for joining 
iso_codes %<>%  rename("Location" = NAME_0)

# join ISO codes to malaria data for mapping 
geoDat <- malaria_data %>% inner_join(iso_codes, by = "Location")

```


Filter the data for 2019 and 2020
```{r}
geoDat2019 <- geoDat %>% filter(Period == 2019)
geoDat2020 <- geoDat %>% filter(Period == 2020)
```

Map log transformed cases in 2019
```{r}
# Create a chloropleth of cases in 2019
figCases2019 <- plot_geo(data = geoDat2019, locations = ~ISO) %>%
  add_trace(
    z = ~log10(clean_cases +1),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "log(Cases) <br>2019") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
    dragmode = F
  )

# Show the plot
figCases2019

```


Map malaria log transformed cases in 2020 
```{r}
figCases2020 <- plot_geo(data = geoDat2020, locations = ~ISO) %>%
  add_trace(
    z = ~ log10(clean_cases +1),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "log(Cases) <br>2020") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
     dragmode = F # disable drag mode with mouse.
  )

# Show the plot
figCases2020

```

Rename total population columns for 2019 and 2020 subset. 
```{r}
geoDat2019 %<>%  rename("Total_Population" = `Total Population`)
geoDat2020 %<>% rename("Total_Population" = `Total Population`)
```

```{r}
# Create a chloropleth of Malaria Prevalence in 2019
figPrev2019 <- plot_geo(data = geoDat2019, locations = ~ISO) %>%
  add_trace(
    z = ~round((clean_cases/Total_Population*100),2),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "Prevelance (%) <br>2019") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
    dragmode = F #disable drag mode with mouse.
  )

# Show the plot
figPrev2019

```

```{r}
# Create a chloropleth of malaria prevalence in 2020
figPrev2020 <- plot_geo(data = geoDat2020, locations = ~ISO) %>%
  add_trace(
    z = ~round((clean_cases/Total_Population)*100,2),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T, 
    zmin = 0,  # Set the minimum value for the color scale
    zmax = 40  # set max value for the color scale 
  ) %>%
  colorbar(title = "Prevelance (%) <br>2020") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
    dragmode = F #disable drag mode with mouse. 
  )

# Show the plot
figPrev2020
```

Examine the log transformed deaths in 2019

```{r}
# Create a chloropleth map of Africa log transformed deaths
figDeaths2019 <- plot_geo(data = geoDat2019, locations = ~ISO) %>%
  add_trace(
    z = ~log10(clean_deaths +1),
    #text = ~paste("Country: ", ISO "<br>Cases: ", clean_cases),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "log(Deaths) <br>2019") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
     dragmode = F
  )

# Show the plot
figDeaths2019
```

Examine the log transformed deaths in 2020
```{r}
fig2020_deaths <- plot_geo(data = geoDat2020, locations = ~ISO) %>%
  add_trace(
    z = ~log10(clean_deaths + 1),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "log(Deaths)  <br>2020") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
     dragmode = F # disable drag mode for mouse 
  )

# Show the plot
fig2020_deaths
```

Map the Malaria Mortality Rate in 2019
```{r}
fig2019Mortality <- plot_geo(data = geoDat2019, locations = ~ISO) %>%
  add_trace(
    z = ~(clean_deaths/Total_Population)*100,
    #text = ~paste("Country: ", ISO "<br>Cases: ", clean_cases),
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "Mortality Rate (%)  <br>2019") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
     dragmode = F
  )

# Show the plot
fig2019Mortality
```
Map the Mortality Rate 2020

```{r}
fig2020Mortality <- plot_geo(data = geoDat2020, locations = ~ISO) %>%
  add_trace(
    z = ~(clean_deaths/Total_Population)*100,
    locations = ~ISO,
    type = 'choropleth',
    colorscale = 'Viridis',
    showscale = T
  ) %>%
  colorbar(title = "Mortality Rate (%)  <br> 2020") %>%
  layout(
    geo = list(
      scope = 'africa'
    ),
     dragmode = F
  )

# Show the plot
fig2020Mortality
```


Create a shiny dashboard

Define the UI so it displays 2019 and 2020 graphs in rows and columns for each cooresponding  epidemiological meesure.
```{r}
ui <- fluidPage(
  # title the dashboard
  titlePanel("Malaria Dashboard"),
  fluidRow(
    column(
      width = 6, # Set the width to half of the row for 2019
      plotlyOutput("plot1"),
      plotlyOutput("plot2"),
      plotlyOutput("plot3"),
      plotlyOutput("plot4")
    ),
    column(
      width = 6, # Set the width to half of the row for 2020
      plotlyOutput("plot5"),
      plotlyOutput("plot6"),
      plotlyOutput("plot7"),
      plotlyOutput("plot8")
    )
  )
)



```

Define the server to host the plotly plots made above. 

```{r}
server <- function(input, output) {
  # Create the cases in 20191
  output$plot1 <- renderPlotly({
    figCases2019
  })
  # deaths in 2019
  output$plot2 <- renderPlotly({
    figDeaths2019
  })
  # 2019 prevalence
  output$plot3 <- renderPlotly({
    figPrev2019
  })
  # 2019 mortality
  output$plot4 <- renderPlotly({
    fig2019Mortality
  })
  
  # Create the Plotly graphs for 2020
  output$plot5 <- renderPlotly({
    figCases2020
  })
  
  output$plot6 <- renderPlotly({
    fig2020_deaths
  })
  
  output$plot7 <- renderPlotly({
    figPrev2020
  })
  
  output$plot8 <- renderPlotly({
    fig2020Mortality
  })
}

# launch the app
shinyApp(ui, server)


```


Calculate the correlations 

```{r}
# obtain the numeric columns to calcualte correlations 
malaria_numeric <- malaria %>%  select(-c("...1", "Location", "Period"))

# calculate Spearman's correlations 
corr <- cor(malaria_numeric, method = "spearman")

```

Create a correlation matrix 
```{r}
ggcorrplot(corr,
  hc.order = T, type = "lower",
  lab = T
)
```

View relationships of all variables. 
```{r}
plot(malaria_numeric)
```

