---
title: "FinalProjectJoins"
author: "Seamus Stein"
date: "11/1/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Purpose: 

The purpose of this markdown is to join all of the tables of malaria case information, deaths, and demographic/census data into one data frame for analysis.

Inputs. The following files are read in:

final_cases: a csv file containing countries and the cleaned cases 
final_deaths: a csv file containing countries and the cleaned deaths data
final_demo: a csv file containing demographic/census information by country. This only contains data from 2023, but used to see what is excluded. 
cleaned_census.csv: a csv file containing demographic/census information by country containing data from 2019-2023.

Output:

final_data.csv: a csv file containing all the data joined. This will be used in the next step for regression analysis and mapping as well as other data visualization. 

```{r}
library(tidyverse)
library(magrittr)
```

Read in the clean data sets
```{r}
final_cases <- read_csv("/Users/seamus/Desktop/project_data/final_cases")
final_deaths <- read_csv("/Users/seamus/Desktop/project_data/final_deaths")
final_demo <- read_csv("/Users/seamus/Desktop/project_data/final_demo")
census_clean <- read_csv("~/Desktop/project_data/cleaned_census.csv")
```


```{r}
head(final_cases)
head(final_deaths)
head(final_demo)
head(census_clean)
```

```{r}
setdiff(final_cases$Location, final_deaths$Location)
```

```{r}
setdiff(final_demo$Name, final_cases$Location)
```

```{r}
setdiff(final_demo$Name, final_deaths$Location)
```

Mauritius is the difference of the set differences

Join the tables
```{r}
#join by location and period at once 
cases_deaths <- final_cases %>% inner_join(final_deaths, by = c("Location" = "Location", "Period" = "Period")) %>% 
                                select(Location, Period, clean_cases, clean_deaths)
      
```

```{r}
"Mauritius" %in% cases_deaths$Location
```
```{r}
setdiff(cases_deaths$Location, final_demo$Name)
```



```{r}
all_data1 <- cases_deaths %>% inner_join(final_demo, by = c("Location" = "Name")) %>% select(-c("...1"))
head(all_data1)
```
```{r}
length(unique(all_data1$Location)) #43 countries
length(unique(final_demo$Name)) # 55
length(unique(cases_deaths$Location)) #47 countries
```
```{r}
"Côte d’Ivoire" %in% all_data1$Location
"Congo" %in% all_data1$Location
```


Join cases and deaths  with census_clean data
```{r}
all_data2 <- cases_deaths %>% inner_join(census_clean, by = c("Location" = "Name", "Period" = "Year"))
head(all_data2)
```


Remove the rows with NAs in clean_cases and or clean_deaths. 


```{r}
final_all_data <- na.omit(all_data2)

# should be now not in countries in the data 
"South Africa" %in% final_all_data$Location
"Eswatini" %in% final_all_data$Location
```

```{r}
"Côte d’Ivoire" %in% all_data2$Location
"Congo" %in% all_data2$Location
```
```{r}
length(unique(all_data2$Location)) #43 countries
length(unique(census_clean$Name)) # 231 countries
length(unique(cases_deaths$Location)) #47 countries
```

Write out the final data for regression analysis. 
```{r}
write.csv(final_all_data, "final_data.csv", row.names=T)
```

